---
title: "Chapter 6"
author: "Ioanna Bouri"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Week 6

## Task 1: Analysis on the RATS data

## Individual response profiles by group for the RATS data.

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)

RATS <- read.csv("data/rats.csv", header=TRUE)

RATS$ID <- factor(RATS$ID)
RATS$Group <- factor(RATS$Group)

# Draw the plot
ggplot(RATS, aes(x = time, y = rats, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATS$rats), max(RATS$rats)), name='Observed weight (grams)')
```


## Individual response profiles for RATS data after standardization.

```{r}
# Standardise the variable rats
RATS <- RATS %>%
  group_by(time) %>%
  mutate(stdrats = (rats - mean(rats))/sd(rats) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATS)

# Plot again with the standardised rats
ggplot(RATS, aes(x = time, y = stdrats, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "standardized weight (grams)")
```

We can observe that both the original and standardized versions are rather close to each other for each group. This indicates that probably there are not many outlying values in the data. This plot essentially represents the rats variable (or its standardized version, respectively) over the time of the study, for each different group.

## Mean response profiles for the three groups in the RATS data.

```{r, echo=FALSE}
# Number of times
n <- RATS$time %>% unique() %>% length()

# Summary data with mean and standard error of bprs by treatment and week 
RATSS <- RATS %>%
  group_by(Group, time) %>%
  summarise( mean = mean(rats), se = sd(rats)/sqrt(n) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSS)

# Plot the mean profiles
ggplot(RATSS, aes(x = time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.8)) +
  scale_y_continuous(name = "mean(weight) +/- se(weight)")
```

We can observe that this plot shows that the behavior of the experiment over time follows a similar pattern for all three groups. Worth mentioning is the difference of the first group, although the pattern is similar with the other two. Also, we can observe that the results from group 2 have greater variance, than what we have observed for groups one and three.

## Boxplots for the RATS data.

```{r, echo=FALSE}
# Create a summary data by Group and ID with mean as the summary variable
RATS8S <- RATS %>%
  filter(time > 0) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(rats) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATS8S)

# Draw a boxplot of the mean versus treatment
ggplot(RATS8S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(weight), time 1-8")
```


The above observations about the different levels of variance become more evident in the boxplots.


## Task 2: Analysis on the BPRS data

## Plot individual BPRS profiles for each treatment group

```{r}
BPRS <- read.csv("data/bprs.csv", header=TRUE)

BPRS <- BPRS %>%
  group_by(week) %>%
  mutate(stdbprs = (bprs - mean(bprs))/sd(bprs) ) %>%
  ungroup()

BPRS$subject <- factor(BPRS$subject)
BPRS$treatment <- factor(BPRS$treatment)

ggplot(BPRS, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRS$bprs), max(BPRS$bprs)))
```

## Fitting the linear model

```{r}
# create a regression model BPRS_reg
BPRS_reg <- lm(bprs ~ week + treatment, data = BPRS)

# print out a summary of the model
summary(BPRS_reg)
```

## Creating a random intercept model
```{r}
# access library lme4
library(lme4)

# Create a random intercept model
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRS, REML = FALSE)

# Print the summary of the model
summary(BPRS_ref)

```

## Random intercept and random slope model
```{r}
# create a random intercept and random slope model
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRS, REML = FALSE)

# print a summary of the model
summary(BPRS_ref1)

# perform an ANOVA test on the two models
anova(BPRS_ref1, BPRS_ref)
```

## Model with interaction
```{r}
# create a random intercept and random slope model
BPRS_ref2 <- lmer(bprs ~ week * treatment + (week | subject), data = BPRS, REML = FALSE)

# print a summary of the model
summary(BPRS_ref2)

# perform an ANOVA test on the two models
anova(BPRS_ref2, BPRS_ref1)

# draw the plot of BPRS
ggplot(BPRS, aes(x = week, y = bprs, group = subject)) +
  #geom_line(aes(linetype = treatment)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2)) +
  geom_point(size=2) +
  scale_shape_manual(values = c(1,2)) +
  scale_x_continuous(name = "Week", breaks = seq(0, 60, 20)) +
  scale_y_continuous(name = "bprs") +
  theme(legend.position = "top")
```


```{r}
# Create a vector of the fitted values
Fitted <- fitted(BPRS_ref2)

# Create a new column fitted to RATSL
BPRS <- BPRS %>%
  mutate(Fitted)

# draw the plot of Fitted BPRS
ggplot(BPRS, aes(x = week, y = Fitted, group = subject)) +
  #geom_line(aes(linetype = treatment)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2)) +
  geom_point(size=2) +
  scale_shape_manual(values = c(1,2)) +
  scale_x_continuous(name = "Week", breaks = seq(0, 60, 20)) +
  scale_y_continuous(name = "bprs") +
  theme(legend.position = "top")
```
